---
title: "Reproducible Reporting using Markdown - R/RStudio"
author: "Meta Miriam Bönniger"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    code_folding: 'hide'
---

# {.tabset}


```{r setup}
knitr::opts_chunk$set(echo = TRUE)
```

**Advanced markdown script example**
    
-----

## Setup

### Load packages
```{r load packages, message=FALSE}
list.of.packages <- c("patchwork", "ggpubr", "tidyverse", "data.table", "knitr", "markdown", "rmarkdown", "dplyr", "ggplot2", "moments", "car", "tableone")

# install packages if new
new.packages <- list.of.packages[!(list.of.packages %in% installed.packages()[,"Package"])]
if(length(new.packages)) install.packages(new.packages)

# load packages
invisible(lapply(list.of.packages, library, character.only = TRUE))
```

### Download OASIS dataset

We'll be downloading the data from <http://www.oasis-brains.org/pdf/oasis_longitudinal.csv> 

A full explanation of the data can be found in: 

*Open Access Series of Imaging Studies (OASIS): Longitudinal MRI Data in Nondemented and Demented Older Adults.*
*Marcus, DS, Fotenos, AF, Csernansky, JG, Morris, JC, Buckner, RL, 2010. Journal of Cognitive Neuroscience, 22, 2677-2684. doi: 10.1162/jocn.2009.21407*  

```{r load data}
df <- read.csv("http://www.oasis-brains.org/pdf/oasis_longitudinal.csv") 

# print out top 5 rows of the DataFrame
kable(head(df, n = 5L)) #if the package kable is not found change it to knitr::kable
```

## Pre-processing the Data

### Dementia status from CDR
We're told in the associated paper (linked above) that the CDR values correspond to the following levels of severity:

* 0 = no dementia
* 0.5 = very mild AD
* 1 = mild AD
* 2 = moderate AD

So let's derive this "Status" from the CDR values:

```{r CDR change}
#create an ordered factor and map CDR values to status 
df <- df %>%
  #leave numeric version for later
  dplyr::mutate(CDR.num = CDR) %>%
  #Normally, mutate works iteratively within a command. However, I have experienced situations 
  #where the newly created variables could not be reused in the same mutate, so here is a second mutate function.
  dplyr::mutate(CDR = dplyr::case_when(CDR.num == 0 ~ 'no dementia',
                                       CDR.num == 0.5 ~ 'very mild AD',
                                       CDR.num == 1 ~ 'mild AD',
                                       CDR.num == 2 ~ 'moderate AD'), # if no cases match NA is returned
                CDR = forcats::fct_reorder(CDR, CDR.num))
```

Now we can look at this status, and how many records there are both overall, and at each of the visits (or time points).

```{r CDR status}
# Let's look at how many records we have of each severity
kable(df %>%
        dplyr::group_by(CDR) %>%
        dplyr::tally())
```

```{r CDR status per visit}
# How many of each status do we have at each visit?
kable(df %>%
        dplyr::group_by(Visit,CDR) %>%
        dplyr::tally())
```

### Variable Types
```{r variable types, comment=NA}
str(df)
```

Change variable names
```{r variable names}
df <- df %>%
  dplyr::rename(Sex = M.F,
                Education = EDUC)
```

**Coerche types? Not always needed but good practice.**

Change "Visit" from numeric values to ordered factor (1-5).
```{r Visit change}
df <- df %>%
  dplyr::mutate(Visit = factor(Visit, 
                               levels=c(1,2,3,4,5),
                               ordered=TRUE))
```

Change "SES" from numeric values to ordered factor (1 = highest to 5 = lowest status).
```{r SES change}
df <- df %>%
  dplyr::mutate(SES = factor(SES, 
                             levels=c(1,2,3,4,5),
                             ordered=TRUE))
```

### Visits & Baseline Data
```{r visit}
kable(df %>%
        dplyr::group_by(Visit) %>%
        dplyr::tally(), 
      caption = "Participant number per visit")
```

Filter for baseline data, and look at their dementia status on arrival
```{r baseline visit}
df_baseline <- df %>% 
  dplyr::filter(Visit == 1)

kable(df_baseline %>%
        dplyr::group_by(CDR) %>%
        dplyr::tally(), 
      caption = "CDR status of participants in first visit")
```

## Data Exploration

### Summarise Data (Non-numerical Variables)
With the exception of CDR, as this has already been done above. 

```{r overview non numerical variables}
numeric.var <- df %>%
  dplyr::select(where(is.numeric), -CDR.num) %>% 
  names()

kable(head(df %>% 
             dplyr::select(-all_of(numeric.var))) %>% 
        data.frame(),
      caption = "Head of all non-numeric variables in the dataset.")

kable(df %>%
        dplyr::group_by(Group) %>%
        dplyr::tally(), 
      caption = "Factor levels and frequency of variable 'Group'")

kable(df %>%
        dplyr::group_by(Visit) %>%
        dplyr::tally(), 
      caption = "Factor levels and frequency of variable 'Visit'")

kable(df %>%
        dplyr::group_by(Sex) %>%
        dplyr::tally(), 
      caption = "Factor levels and frequency of variable 'Sex'")

kable(df %>%
        dplyr::group_by(Hand) %>%
        dplyr::tally(), 
      caption = "Factor levels and frequency of variable 'Hand'")

kable(df %>%
        dplyr::group_by(SES) %>%
        dplyr::tally(), 
      caption = "Factor levels and frequency of variable 'SES'")
```

### Summarise Data (Numerical Variables)
```{r overview numerical}
kable(df %>% 
        dplyr::select(all_of(numeric.var)) %>%
        summary() %>% 
        data.frame() %>%
        dplyr::select(-Var1) %>%
        dplyr::rename(Variable = Var2) %>%
        dplyr::mutate(Freq = gsub(" ", "", Freq, fixed = TRUE),
                      Freq = gsub("\'", "", Freq, fixed = TRUE)) %>%
        dplyr::filter(!is.na(Freq)) %>%
        tidyr::separate(Freq, c("V1", "V2"), sep= ":") %>%
        dplyr::group_by(Variable) %>%
        tidyr::spread(V1, V2),
      caption = "Summary of all numeric variables in the dataset")
```

### Visualize Data (numerical variables)
```{r histograms, out.width="50%"}
par(mar = c(4, 4, 1, .1))
#distribution
distribution_plots <- function(variable) {
  df.plot <- df %>% 
    dplyr::filter(!is.na(.[[variable]])) 
  
  hist(df.plot[[variable]], 
       breaks = 50, 
       xlab = paste0(variable),
       main = paste0("Histogram of ", variable))
}

invisible(lapply(numeric.var, distribution_plots))

rm(distribution_plots)
```

```{r stratified histograms, message = FALSE, out.width="50%"}
par(mar = c(4, 4, 1, .1))
numeric.var <- c("Age", "Education", "MMSE")
#distribution
distribution_plots <- function(variable) {
  df.plot <- df %>% 
    dplyr::filter(!is.na(.[[variable]])) 
  
  plot <- ggplot(data = df.plot, 
                 aes_(x = as.name(variable), fill = ~ CDR)) +
    geom_histogram()
  print(plot)
}

invisible(lapply(numeric.var, distribution_plots))

rm(numeric.var)
rm(distribution_plots)
```

## Sample Analysis: Table 1

Create a typical table one. Show it in this Markdown and save it also as csv for integrating it, for example, into a manuscript. 

```{r table 1}
# 1. Select variables for table
myVars <- names(df %>%
                  dplyr::select(-Subject.ID, -MRI.ID, -CDR.num))


# 2. Make a list of all categorical variables from the myVars list for this table
catVars <- df %>% 
  dplyr::select(all_of(myVars)) %>%
  dplyr::select_if(purrr::negate(is.numeric)) %>% 
  names()

# 3. Check visually for non-normally distributed numerical variables 
#hist() or histograms
nonnormaldist <- c("MR.Delay", "MMSE")

# 4. Create dataset for table 1
table1df <- df %>%
  dplyr::select(all_of(myVars))

# 5. Create table
tab1 <- tableone::CreateTableOne(data = table1df, 
                                 vars = myVars, 
                                 factorVars = catVars)

# 6. Modify list of table to get a data frame
tab2 <- print(tab1, 
              showAllLevels = TRUE, 
              printToggle = FALSE, 
              noSpaces = TRUE, 
              explain = FALSE, 
              test = TRUE, 
              contDigits = 1, 
              catDigits = 1, 
              missing = TRUE,
              nonnormal = nonnormaldist) %>%
  as.data.frame() %>% 
  data.table::setDT(., keep.rownames = TRUE)
#summary(tab1) #gives more information regarding missing values

# 7. Rename column and rownames (of rn) to create a nice table that can be used in a manuscript
tab3 <- tab2 %>%
  dplyr::mutate(
    rn = ifelse(grepl("^X", rn), "", rn)) %>%
  dplyr::mutate(
    rn = dplyr::recode(rn,
                       "Group" = "Group, N (%)",
                       "Visit" = "Number of participants per visit, N (%)",
                       "MR.Delay" = "Time delay for MR imaging, median [IQR]",
                       "Sex" = "Sex, N (%)",
                       "Hand" = "Handedness, N (%)",
                       "Age" = "Age (years), M (SD)",
                       "Education" = "Education (years), M (SD)",
                       "SES" = "Socioeconomic status, N (%)",
                       "MMSE" = "Mini mental state examination, median [IQR]",
                       "CDR" = "Clinical Dementia Rating, N (%)",
                       "eTIV" = "Estimated total intracranial volume (cm3), M (SD)",
                       "nWBV" = "Normalized whole-brain volume, M (SD)",
                       "ASF" = "Atlas scaling factor, M (SD)"),
    level = dplyr::recode(level,
                          "F" = "women",
                          "M" = "men",
                          "R" = "right")) %>%
  dplyr::rename("Characteristic" = "rn") 

# 8. Print table in markdown
kable(tab3)

# 9. Save table for manuscript
write.csv(tab3, file = "Table_1.csv", row.names = FALSE) 

rm(tab1); rm(tab2); rm(tab3); rm(table1df); rm(catVars); rm(myVars); rm(nonnormaldist)
```

## Data Analysis: MMSE Trajectories over Time
```{r longitudinal associations}
df %>% 
  dplyr::select(Subject.ID, MR.Delay, MMSE, CDR) %>%
  dplyr::filter(complete.cases(.)) %>%
  ggplot(aes(x = MR.Delay, y = MMSE, color = CDR, group = Subject.ID)) +
  geom_line()
```

## Data Analysis: Regression Analysis

### Simple Linear Regression
```{r simple linear regression, comment = NA, message=FALSE}
# To be able to create a multitude of plots with different y-axis values I created a function for the plots and used a loop (lapply) to create them. 
# We have to add numeric sex and CDR variables to get regression lines in ggplot
df <- df %>%
  dplyr::mutate(Sex.num = ifelse(Sex == "M", 1, 
                                 ifelse(Sex == "F", 2,
                                        NA)))

plotAgeAssociation <- function(xaxis) {
  # create complete cases dataset
  df.plot <- df %>%
    dplyr::select(MMSE, as.name(xaxis), CDR) %>%
    filter(complete.cases(.))
  
  print(ggplot(data=df.plot, aes_(x=as.name(xaxis), y=~MMSE)) + 
          geom_point(aes_(color=~CDR)) +
          geom_smooth(method = lm) +
          theme(legend.position = "bottom"))
  regr <- lm(as.formula(paste0("MMSE ~ ", xaxis)), 
             data = df.plot %>%
               # We recode CDR back as an unordered factor to have more interpretable results.
               dplyr::mutate(CDR = factor(CDR, ordered=FALSE)))
  result <- broom::tidy(regr, conf.int = TRUE, conf.level = 0.95)
  result$N <- regr$residuals %>% length()
  result$r.squared <- summary(regr)$r.squared
  result <- result %>%
    dplyr::mutate(across(where(is.numeric), round, 3))
  print(kable(result))
}

# run functions for education and MMSE on the y-axis
invisible(lapply(c("Age", "Sex.num", "Education", "CDR.num"), plotAgeAssociation))
rm(plotAgeAssociation)
```

**Then we can do the same again, but this time stratify by the dementia status (not including CDR this time):**
```{r simple linear regression CDR strat, message = FALSE, out.width="50%"}
par(mar = c(4, 4, 1, .1))
plotAgeAssociation <- function(xaxis) {
  # create complete cases dataset
  df.plot <- df %>%
    dplyr::select(MMSE, as.name(xaxis), CDR) %>%
    dplyr::filter(complete.cases(.))
  
  print(ggplot(data=df.plot, aes_(x=as.name(xaxis), y=~MMSE, color=~CDR)) + 
          geom_point() +
          geom_smooth(method = lm, se = FALSE) +
          theme(legend.position = "bottom"))
}

# run functions for education and MMSE on the y-axis
invisible(lapply(c("Age", "Sex.num", "Education"), plotAgeAssociation))
rm(plotAgeAssociation)
```

### Multiple Linear Regression

We create a regression model with MMSE as the outcome and age, sex, education and CDR as the independents. Use the no dementia group as reference group for CDR.

```{r regression model}
model <- lm(MMSE ~ Age + Sex + Education + relevel(factor(CDR, ordered=FALSE), ref = "no dementia"), 
            data = df_baseline)
```

#### Assumption check
```{r skewness of residuals and vif check}
kable(moments::skewness(model$residuals) %>%
        data.frame() %>%
        dplyr::rename("Skewness of residuals" = "."), 
      caption = "Skewness of model residuals")
kable(car::vif(model), 
      caption = "Variance-inflation factors to assess multicolinearity of model independents")
```

Plots to check other regression model assumptions
```{r plots to check the regression assumptions}
par(mfrow = c(2,2)) # Change the panel layout to 2 x 2
plot(model)
```

#### Result table
```{r regression model results}
# Creating data frame including all results
result<- broom::tidy(model, conf.int = TRUE, conf.level = 0.95)
result$N <- model$residuals %>% length()
result$r.squared <- summary(model)$r.squared

# Modify outcome for export
result %>%
  dplyr::rename(independent = term) %>%
  dplyr::mutate(
    independent = dplyr::recode(independent,
                                SexM = 'Sex (ref. women)',
                                Education = 'Education (in years)',
                                'relevel(factor(CDR, ordered = FALSE), ref = \"no dementia\")mild AD' = 'CDR (ref. no dementia) mild AD',
                                'relevel(factor(CDR, ordered = FALSE), ref = \"no dementia\")very mild AD' = 'CDR (ref. no dementia) very mild AD')) %>%
  dplyr::mutate(across(where(is.numeric), round, 3)) %>%
  kable(., caption = "Outcome: MMSE")
```

## References R session information
```{r session info, comment=NA}
citation()
sessionInfo() 
lapply(list.of.packages, citation)

#RStudio.Version()[-2] #both functions are not working in 
#rstudioapi::versionInfo()[-2] 
## both functions for the R Studio version are not working in R markdown. Run this code in the console and enter information manually. 
```

To cite RStudio in publications use:
RStudio Team (2019). RStudio: Integrated Development for R. RStudio, Inc., Boston, MA URL <http://www.rstudio.com/.version>
Version: 1.2.5033
Release name: "Orange Blossom"




